Pareto-Optimal Multistate Protein Design using Tree Graphical Models

### 2012-12-06 docs/tex/draft.tex
Added the meat of how to max decoding on Tree MRFs.

\DeclareMathOperator*{\argmax}{arg\,max}

@@

\begin{align*}
X_A &= \arg \min_X -\log  \left[{\prod_{i=1}^{n}{\phi(X_i)\phi(\mathbf{O^A_i})\psi(\mathbf{O^A_i},X_i})\prod_{i=1}^{n-1}{\psi(X_i,X_{i+1})}}\right]
\end{align*}

\begin{figure}[h!]
% selecting unit length
\centering
% used for centering Figure
\setlength{\unitlength}{0.3cm}
\begin{picture}(40,25)(0,0)
\multiput(5,15)(8,0){2}{\circle{4}}
\multiput(5,7)(8,0){2}{\circle{4}}
\multiput(5,7)(8,0){2}{\circle{3.7}}
\multiput(5,23)(8,0){2}{\circle{4}}
\multiput(5,23)(8,0){2}{\circle{3.7}}
\thicklines
\multiput(7,15)(8,0){2}{\line(1,0){4}}
\multiput(20,15)(1,0){4}{\circle*{0.3}}
\multiput(5,13)(8,0){2}{\line(0,-1){4}}
\multiput(5,17)(8,0){2}{\line(0,1){4}}
\put(24,15){\line(1,0){4}}
\put(30,13){\line(0,-1){4}}
\put(30,17){\line(0,1){4}}
\thinlines
\put(30,15){\circle{4}}
\put(30,7){\circle{4}}
\put(30,7){\circle{3.7}}
\put(30,23){\circle{4}}
\put(30,23){\circle{3.7}}
\put(3.5,14.6){$\phi(X_1)$}
\put(11.5,14.6){$\phi(X_2)$}
\put(28.5,14.6){$\phi(X_n)$}
\put(3.5,6.6){$\phi(O^A_1)$}
\put(11.5,6.6){$\phi(O^A_2)$}
\put(28.5,6.6){$\phi(O^A_n)$}
\put(3.5,22.6){$\phi(O^B_1)$}
\put(11.5,22.6){$\phi(O^B_2)$}
\put(28.5,22.6){$\phi(O^B_n)$}
\put(7,16.5){\footnotesize{$\psi(X_1{,}X_2)$}}
\put(15,16.5){\footnotesize{$\psi(X_2{,}X_3)$}}
\put(22.5,16.5){\footnotesize{$\psi(X_{n{-}1}{,}X_n)$}}
\put(5.2,10.6){\footnotesize{$\psi(X_1{,}O^A_1)$}}
\put(13.2,10.6){\footnotesize{$\psi(X_2{,}O^A_2)$}}
\put(30.2,10.6){\footnotesize{$\psi(X_n{,}O^A_n)$}}
\put(5.2,18.6){\footnotesize{$\psi(X_1{,}O^B_1)$}}
\put(13.2,18.6){\footnotesize{$\psi(X_2{,}O^B_2)$}}
\put(30.2,18.6){\footnotesize{$\psi(X_n{,}O^B_n)$}}
\end{picture}
\label{fig:CMRF2}
\caption{A Tree structured Markov Random Field that is used for modelling multi-state protein design.}
\end{figure}

@@

\subsection{Convex Hull Procedure on CMRFs}

>>>>

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

@@

\begin{align*}
X_A &= \arg \min_X E_A(X) \\
&= \arg \min_X -\log  \left[{\prod_{i=1}^{n}{\phi(X_i)\phi(\mathbf{O^A_i})\psi(\mathbf{O^A_i},X_i})\prod_{i=1}^{n-1}{\psi(X_i,X_{i+1})}}\right]\\
&= \arg \min_X -\left[{\sum_{i=1}^{n} \log\phi(X_i)+\sum_{i=1}^{n} \log\phi(\mathbf{O^A_i})+\sum_{i=1}^{n} \log\psi(\mathbf{O^A_i},X_i)+ \sum_{i=1}^{n-1}{\log\psi(X_i,X_{i+1})}}\right] \\
&= \arg \min_X -\left[{\sum_{i=1}^{n} \log\phi(X_i)+\sum_{i=1}^{n} \log\psi(\mathbf{O^A_i},X_i)+ \sum_{i=1}^{n-1}{\log\psi(X_i,X_{i+1})}}\right]
\end{align*}

Similarly,
\begin{align*}
X_B &= \arg \min_X E_B(X) \\
&= \arg \min_X -\left[{\sum_{i=1}^{n} \log\phi(X_i)+\sum_{i=1}^{n} \log\psi(\mathbf{O^B_i},X_i)+ \sum_{i=1}^{n-1}{\log\psi(X_i,X_{i+1})}}\right]
\end{align*}
\\
Taking the convex combination of the energy functions we get 
\begin{align*}
X_{AB} &= \arg \min_X \theta_A E_A(X) + \theta_B E_B(X) \\
\text{s.t} &\quad  \theta_A + \theta_B  = 1 \\
& \theta_A, \theta_B \geq 0
\end{align*}
Expanding $E_A(X)$ and $E_B(X)$ we get, 
\begin{align*}
X_{AB} &= \arg \min_X -\theta_A \left[{\sum_{i=1}^{n} \log\phi(X_i)+\sum_{i=1}^{n} \log\psi(\mathbf{O^A_i},X_i)+ \sum_{i=1}^{n-1}{\log\psi(X_i,X_{i+1})}}\right] \\
&-\theta_B \left[{\sum_{i=1}^{n} \log\phi(X_i)+\sum_{i=1}^{n} \log\psi(\mathbf{O^B_i},X_i)+ \sum_{i=1}^{n-1}{\log\psi(X_i,X_{i+1})}}\right] \\
&= \arg \min_X -\left( \sum_{i=1}^{n}{\log\phi(X_i)} + \sum_{i=1}^{n-1}{\log\psi(X_i,X_{i+1})}\right)\\ 
 &-\left( \sum_{i=1}^{n}{(\theta_A\log\psi(\mathbf{O^A_i},X_i) + \theta_B\log\psi(\mathbf{O^B_i},X_i))} \right) \\
 &= \arg \min_X -\left( \sum_{i=1}^{n}{\log\phi(X_i)} + \sum_{i=1}^{n-1}{\log\psi(X_i,X_{i+1})}\right)\\ 
 &-\left( \sum_{i=1}^{n}{(\log\psi(\mathbf{O^A_i},X_i)^{\theta_A} + \log\psi(\mathbf{O^B_i},X_i)^{\theta_B})} \right)
\end{align*}

This looks very much like decoding on a CMRF with the added change that the emission edge potential values are exponentiated. In fact this distribution factorizes according to the TMRF shown in Figure~\ref{fig:CMRF2}.

\begin{figure}[h!]
% selecting unit length
\centering
% used for centering Figure
\setlength{\unitlength}{0.3cm}
\begin{picture}(40,25)(0,0)
\multiput(5,15)(8,0){2}{\circle{4}}
\multiput(5,7)(8,0){2}{\circle{4}}
\multiput(5,7)(8,0){2}{\circle{3.7}}
\multiput(5,23)(8,0){2}{\circle{4}}
\multiput(5,23)(8,0){2}{\circle{3.7}}
\thicklines
\multiput(7,15)(8,0){2}{\line(1,0){4}}
\multiput(20,15)(1,0){4}{\circle*{0.3}}
\multiput(5,13)(8,0){2}{\line(0,-1){4}}
\multiput(5,17)(8,0){2}{\line(0,1){4}}
\put(24,15){\line(1,0){4}}
\put(30,13){\line(0,-1){4}}
\put(30,17){\line(0,1){4}}
\thinlines
\put(30,15){\circle{4}}
\put(30,7){\circle{4}}
\put(30,7){\circle{3.7}}
\put(30,23){\circle{4}}
\put(30,23){\circle{3.7}}
\put(3.5,14.6){$\phi(X_1)$}
\put(11.5,14.6){$\phi(X_2)$}
\put(28.5,14.6){$\phi(X_n)$}
\put(3.5,6.6){$\phi(O^A_1)$}
\put(11.5,6.6){$\phi(O^A_2)$}
\put(28.5,6.6){$\phi(O^A_n)$}
\put(3.5,22.6){$\phi(O^B_1)$}
\put(11.5,22.6){$\phi(O^B_2)$}
\put(28.5,22.6){$\phi(O^B_n)$}
\put(7,16.5){\footnotesize{$\psi(X_1{,}X_2)$}}
\put(15,16.5){\footnotesize{$\psi(X_2{,}X_3)$}}
\put(22.5,16.5){\footnotesize{$\psi(X_{n{-}1}{,}X_n)$}}
\put(5.2,10.6){\footnotesize{$\psi^{\theta_A}(X_1{,}O^A_1)$}}
\put(13.2,10.6){\footnotesize{$\psi^{\theta_A}(X_2{,}O^A_2)$}}
\put(30.2,10.6){\footnotesize{$\psi^{\theta_A}(X_n{,}O^A_n)$}}
\put(5.2,18.6){\footnotesize{$\psi^{\theta_B}(X_1{,}O^B_1)$}}
\put(13.2,18.6){\footnotesize{$\psi^{\theta_B}(X_2{,}O^B_2)$}}
\put(30.2,18.6){\footnotesize{$\psi^{\theta_B}(X_n{,}O^B_n)$}}
\end{picture}
\label{fig:CMRF2}
\caption{A Tree structured Markov Random Field that is used for modelling multi-state protein design. This Tree structured MRF results from taking a convex combination $\theta_A E_A(X) + \theta_B E_B(X)$ of the log likelihoods of CMRFs for State A and State B }
\end{figure}

@@

\subsection{Max-Decoding on TMRFs}
In the previous section we showed how the convex combination of negative log likelihoods of CMRFs can be used to construct a TMRF. Here we will show how efficient decoding can be performed on the TMRF using the similar variant of Viterbi decoding.

\begin{align*}
X_{AB} &= \argmin_X \left( \theta_A E_A(X) + \theta_B E_B(X)\right) \\
&= \arg \min_X \left( -\sum_{i=1}^{n}{\log\phi(X_i)} + \sum_{i=1}^{n-1}{\log\psi(X_i,X_{i+1})}\right)\\ 
 &-\left( \sum_{i=1}^{n}{(\log\psi(\mathbf{O^A_i},X_i)^{\theta_A} + \log\psi(\mathbf{O^B_i},X_i)^{\theta_B})} \right) \\
 &= \argmax_{X}\log\left( \prod_{i=1}^{n}{\phi(X_i)\psi(\mathbf{O^A_i},X_i)^{\theta_A} \psi(\mathbf{O^B_i},X_i)^{\theta_B}}\prod_{i=1}^{n-1}{\psi(X_i,X_{i+1})}\right) \\
 &= \argmax_{X}\left( \prod_{i=1}^{n}{\phi(X_i)\psi(\mathbf{O^A_i},X_i)^{\theta_A} \psi(\mathbf{O^B_i},X_i)^{\theta_B}}\prod_{i=1}^{n-1}{\psi(X_i,X_{i+1})}\right)
\end{align*}

\textit{Calculation of the Optimal Probability value:}
\[
  V(i,k) = \left\{
  \begin{array}{l l}
    \psi(\mathbf{O^A_1},X_1{=}k)^{\theta_A}\psi(\mathbf{O^B_1},X_1{=}k)^{\theta_B}\phi(X_1{=}k) & \quad \text{if $i=1$}\\
    \phi(X_i{=}k)\psi(\mathbf{O^A_i},X_i{=}k)^{\theta_A}\psi(\mathbf{O^B_1},X_1{=}k)^{\theta_B}\cdots\\
    \max_{j}\{\psi(X_{i}{=}k,X_{i-1}{=}j)V(i-1,j)\} & \quad \text{if $2\leq i\leq n$}\\
  \end{array} \right.
\]

Here, $V(i,k)$ corresponds to the maximum functional value attainable by examining the sequence from $X_{1:i}$ and the observations from $\mathbf{O^A_{1:i}}$ and $\mathbf{O^B_{1:i}}$. Also, $X^\star_{1:n}$ corresponds to the optimal sequence assignment. Pointers are stored to each of the previous variables during the calculation of $V(i,k)$. These pointers are used to retrieve the optimal sequence assignment.

\subsection{Convex Hull Procedure on TMRFs}
In the previous section we showed how to find $X_{AB} = \argmin_X \left( \theta_A E_A(X) + \theta_B E_B(X)\right)$, for any convex combination with $\theta_A$ and $\theta_B$.In this section, we will put it all together to define a comprehensive algorithm that can be used for finding the entire pareto frontier for sequence design tasks.
